---
title: "RNA Seq miniproject"
author: "Trent Gutierrez"
date: "2022-09-15"
output: 
  html_document:
    theme: "lumen"
    toc: true
    toc_float: 
      collapsed: true
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction

### Research question
This RNA-Seq data investigates genes from neurons differentiated from induced pluripotent stem cells (iPSCs). the **patient** group includes individuals who have either 2q11.2 Deletion Syndrome (22q11.2 DS) or schizoaffective disorder (SAD), this means these individuals are at high risk to developing schizophrenia. This data includes these patient's RNA-Seq data and multiple controls. This R markdown will use the rse data provided by the respective scientist and will conduct analysis to make our own conclusion based off the data. 





### Data Input

This code is just loading all the package libraries we would need for this R markdown. 
```{r library loading, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(BiocManager)
library(recount)
library(RColorBrewer)
library(pheatmap)
library(DT)
library(DESeq2)
library(fgsea)
library(dplyr)
library(ggplot2)
library(plotly)
library(styler)
library(kableExtra)
library(fgsea)
library(stats)
library(clusterProfiler)
library(enrichplot)
library(pathview)
library(org.Hs.eg.db)
library(ReactomePA)
library(reactome.db)
library(devtools)
library(conflicted)
library(stringr)
library(fgsea)
library(data.table)
library(ggplot2)
library(rebus)
library(europepmc)
```


This code was provided by recount2 and was adjusted to download the rse data for the **SRP022028** experiment. The code downloads the `rse_gene.Rdata` file into a folder named "SRP022028" placed into the working directory. the `rse_gene` data is scaled into  `rse_gene.Rdata`   
```{r data input, message=FALSE, warning=FALSE}
url <- download_study("SRP022028") 
load(file.path("SRP022028", 'rse_gene.Rdata'))
```


We must now develop a metadata variable, first we just collect all the colData into the metadata variable. Then we add the condition column which is an identifier adjusted from an existing column. The genotype helps Identify the genetic variable being analyzed. the grepl function looks at the column for the string and returns TRUE or FALSE, and then with a pipe and ifelse, we can say turn the TRUE values into "22q11Del" and FALSE values into "WT"(Wild Type) 
```{r condition setup }

metadata <- colData(rse_gene)

metadata$conditions <- sapply(metadata$characteristics, function(x){ x[1]})

metadata$genotype <- grepl("22q11", metadata$title) %>%
  ifelse("22q11Del","WT")
```

The way we found out the identifiers like `project` and `conditions` was identified by using `colData(rse)` to look at all our columns and try to identify what sample belongs to which condition. 
```{r}
metadata %>% 
  as.data.frame() %>%
  dplyr::select(genotype,conditions) %>%
  kbl(caption = "Metadata for SRP022028") %>%
  kable_styling(bootstrap_options = "striped", full_width = T, html_font = "Cambria")
```


To allow further statistical analysis the `colData()` function was used to view information in `rse` and we can see with the "characteristics" column where our *patient* and *control* samples are indicated. This leads to the addition of the `conditions` column into `rse` so the `DESeq2()` package can interpret the data. The next step is to turn the `rse` data into `dds` which stands for *DESeqDataSet* and then this dataset is ran through `DESeq()` to run the actual differential expression analysis. 
```{r DESeq dataset start,message=FALSE, warning=FALSE}
rse$condition <- conditions

dds <- DESeqDataSet(rse, design = ~condition)
dds <-DESeq(dds)
```


Once the differential expression analysis is complete we still have the `dds` but now it has a **results** function which is assigned to the variable `result` (for later graphs, `result` is also transformed into the data frame `result_dataframe` to make interpretation easier) This `result` variable holds very important information such as the **log2 fold change** and **wald test p-value** using the condition specified earlier *patient* vs *control*



the contrast functions specifies our (condition, numerator, demomenator) this affects the log2basechange value. 
```{r results from the dds}
result <- results(dds, contrast = c("condition","patient","control"))
result_dataframe <- as.data.frame(result)
```


## Results


### Signifcant Differentially Expressed Genes (DEGs)

Now we can start to explore our data, first we want to acquire the actually significantly DEGs from our `result` function. First step is to order the result section by the *p-value*. Next is to filter the results slightly by removing any genes that have a *p-adjusted* value higher than 0.05 (5%) we choose this value likely because it leaves just the genes that gives us around a 95% confidence of being actually significant. To visualize our selected significant DEGs a data table was constructed.      
```{r signifcant DEGs}
orderedresults <- result[order(result$pvalue),]

sigresults <- subset(orderedresults, padj < 0.05)

sigresults_dataframe <- as.data.frame(sigresults)
datatable(sigresults_dataframe, class = 'cell-border stripe' , 
          colnames = c("ENSEMBL Name" = 1), 
          caption = "Table 1: Patient vs Control Significant Genes")
```


### PCA Plot

A PCA plot is used to show how our samples cluster and if it correlates to the treatments. First we want to assign a new variable `vsd_dds` that just normalizes the count from the `dds` data. With the normalized data we construct our PCA graph with the groups *condition* and *sample* to help us identify which samples are which. 
```{r PCA plot}
vsd_dds <- vst(dds)

plotPCA(vsd_dds, intgroup = c("condition","sample"))
```

### Volcano Plot

Volcano plots give a bigger picture view of our data. First we add a new column *threshold* which returns TRUE/FALSE/NA depending if the row has a *p-adjusted* value less than 5% and this is assigned a new variable `result_dataframe_sig_marked`. This data is plotted. 
```{r Volcano plot, message=FALSE, warning=FALSE}
result_dataframe_sig_marked <- data.frame(result_dataframe) %>%
  mutate(threshold = padj < 0.05)

ggplot(result_dataframe_sig_marked) + 
  geom_point(aes(x = log2FoldChange, y= -log10(padj), col = threshold)) + 
  xlim(c(-10,10))+
  ylim(c(0,15))+
  labs(
    x = "log2 Fold Change" , 
    y = "-log10 adjusted p-value" , 
    title = "Volcano plot of signifcant DEGs") +
  theme(legend.position = "bottom", plot.title = element_text(size = rel(1.5), hjust = 0.5), 
              axis.title = element_text(size = rel(1.25)))
```


### Heatmap of top 10 overexpressed & underexpressed DEGs

This graph involves a few data transformations lets go through them step by step. Our first job is to select both the *10 over-expressed* and *10 under-expressed* significant DEGs and then combine them into one dataset called `combinedDEGs`. Next we make sure our counts from `dds` are normalized and added to a new variable `normalized_dds_counts`. Then using the row names from `combinedDEGs` we filter out the normalized count data so we have a variable `sig_norm_dds_counts` with our 20 specific genes, and then that data is heatmapped. `conflict_prefer` is added here just to avoid synthax errors with the `select()` function. 
```{r heatmap}
top10DEGs <- sigresults[1:10,]
bot10DEGs <- sigresults[201:210,]
combinedDEGs <- rbind(top10DEGs,bot10DEGs)

normalized_dds_counts <- counts(dds, normalized=TRUE)
sig_norm_dds_counts <- normalized_dds_counts[rownames(combinedDEGs), ]

heat_colors <- brewer.pal(n = 6, name = "YlOrRd")
conflict_prefer("select","dplyr")

pheatmap(sig_norm_dds_counts, 
         color = heat_colors, 
         cluster_rows = TRUE, 
         show_rownames = TRUE,
         annotation =  select(metadata,conditions), 
         scale = "row")
```


### GSEA

#### Clusterprofiler data set up

The first step of this data set up is to get our `EnsemblIDs` variable, this consist of the row names from the `result_dataframe` however most of them end with a decimal point and value, this causes errors so we use `str_replace` to get rid of the "." and everything pass that point and replace it with nothing so our IDs are in the correct format for gene identification. Now we have to set up our gene list for the clusterprofiler, first it pulls the raw numbers from the result data frame for *log2FoldChange* then the Ensembl IDs are attached to this list. Then any *NA* values are removed, and finally the gene list is ordred from greatest to least, this is neccessary for the `gseGO` function to work correctly. 
```{r Clusterprofiler set up as backup}

EnsemblIDs <- rownames(result_dataframe)
EnsemblIDs <- str_replace(EnsemblIDs,
                           pattern = "\\..*", 
                           replacement = "")

original_gene_list <- result_dataframe$log2FoldChange
names(original_gene_list) <- EnsemblIDs
gene_list <- na.omit(original_gene_list)
gene_list <- sort(gene_list, decreasing = TRUE)

```



`gseGO()` stands for *Gene Set Enrichment Analysis of Gene Ontology* This function will do our enrichment analysis. the gene list is the one we made above, the ont value allows you to analyze for Biological Processes, Molecular Function, and Cellular components, and if you want to see all 3 choose "ALL" like we did here. KeyType is dependent on the IDs we assigned our Gene list earlier, in this dataset we have Ensembl IDs. min and max GSSize gives us the minimum and maximum of the gene count in the list to be analyzed. the p-value cutoff is what the minimum p-value cut off is. P-Adjust method has a variety of different p-value adjusting techniques but for this data we want none of them. (Warning: The Data generated here is larger than anything else in this analysis)
```{r Gene mapped, message=FALSE, warning=FALSE}
gse <- gseGO(geneList = gene_list , ont = "ALL",
             keyType = "ENSEMBL",
             minGSSize = 3, maxGSSize = 800,
             pvalueCutoff = 0.05, verbose = TRUE,
             OrgDb= "org.Hs.eg.db", pAdjustMethod = "none" )
```


#### gseaplot 2 graph

This graph is the top 5 pathways are visualized with the `gseaplot2()` function
```{r GSEA2 Plot}
GSEAplot <- gse %>%
  arrange(p.adjust)%>%
  gseaplot2(geneSetID = 1:5)
GSEAplot
```

## Discussion

Now that all the graphs have been created using our SRP022028 data, we can begin to look at the data generated and begin to make inferences of our samples. All inferences will be based purely on the data generated above.  

The first data to talk about is the identification of significant differentially expressed genes (DEGs) our filtered table has a total of 210 DEGs from a total of 58,037 genes with a 95% confidence of being significant. 

The PCA plot is used to give a good general idea of how our samples cluster. Looking at the plot there is some clustering with the control and patient samples, Patient sample: SRS417724 and SRS417725 are closely clustered together, and another cluster group consist of Patient samples: SRS417723 and SRS417726 with 1 control sample included close by control: SRS417727, and the final cluster group could be the control samples: SRS417728, SRS417730, and SRS417729. The implications from this plot can be that the patient and control samples are reasonably clustered into different sections so the DEGs should actually be somewhat differentially  expressed. 

The Volcano plot shows us all our genes plotted by their log2 Fold change and -log10 p-adj value. The points plotted in blue means that they met the threshold of having a p-adj value lower than 5% giving us our standard 95% confidence. The points in red mean they did not meet the threshold. 

The heatmap of the top 10 over-expressed and 10 under-expressed DEGs shows us the expression of our 8 samples, the top of the heatmap shows us the conditions of the samples on the bottom. The first thing we can see is that the control side has a greater average expression level for most of the genes than the patient side. A possible reason this may be is that with the patient samples that are affected by 22q11.2 DS could have lower expression in these genes depending on where the genes are located and what their function are. 

The gseaplot2 graph shows the enrichment scores for the top 5 pathways sorted by their p-adjusted value. the 5 selected pathways seem to have a very similar enrichment score lines, The peak of this line is what we want to look at. Our graph peaks around rank 37000-38000, and the bottom graph shows our gene's ranked list metric in the negatives which could mean the correlation with our genes can either mean a second phenotype expression or an inverse correlation with their function. 
 
## Conclusion











